Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-07-22 paper.qmd
Update Figure 6
paper.qmd

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: 
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional ($p\text{-}D$) data using non-linear transformation. The methods and parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of $p\text{-}D$ distributions. To help evaluate the NLDR we have developed an algorithm to show the $2\text{-}D$ NLDR model in the $p\text{-}D$ space, viewed with a tour. One can see if the model fits everywhere or better in some subspaces, or completely mismatches the data. It is used to evaluate which $2\text{-}D$ layout is the best representation of the $p\text{-}D$ distribution and see how different methods may have similar summaries or quirks.

keywords: [high-dimensional data, dimension reduction, hexagon binning, low-dimensional representation, tour, data vizualization, model-in-the-data-space]
keywords-formatted: [high-dimensional data, dimension reduction, hexagon binning, low-dimensional representation, tour, data vizualization, model-in-the-data-space]

@@

sc_ltr_pos <- c(0.08, 0.93)
# sc_xlims <- c(-0.5, 1.47)
# sc_ylims <- c(-0.32, 2.1)
sc_xlims <- c(-0.25, 1.25)
sc_ylims <- c(-0.2, 2)

@@

Three views of the true model in projections from $7\text{-}D$, for the S-curve data. The model closely fits the shape. (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/I5GL23vLiw0>). 
:::

<!-- XXX The points on the "theoretical model" should be equidistant along the S. They are still to bunched in some spots and too far apart in others. -->

@@

XXXAdd best fit with UMAP and discuss why it's the best fit

<!-- langevitour with pD model for S-curve -->
::: {#fig-scurve-sc layout-ncol="4" fig-pos="H"}
![](figures/scurve/umap_trimesh_layout.png)

![](figures/scurve/sc_umap_best_1.png)

![](figures/scurve/sc_umap_best_2.png)

![](figures/scurve/sc_umap_best_3.png)

Model in \gD{}, on the UMAP layout, and three views of the fit in projections from $7\text{-}D$, for the S-curve data ($(s_1, \ s_2) = (-0.160, \ -0.263)$, $b = 405 \  (15, \ 27)$, $m = 70$, benchmark value to remove low density hexagons is $0.250$, and benchmark value to remove large edges is $0.159$). MSE is $0.0462$. The model closely fits the shape, but it doesn't fully fill out the width of the S, which means that it does not adequately capture the surface (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/i7F7xpN1Hz8>).
:::

<!-- XXX You need to have consistent colouring. In langevitour the points corresponding to observations are purple. They should be purple in the \gD{} view too.  -->

@@

hex_grid_coloured_scurve1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve1, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  geom_point(data = umap_scurve_scaled,
             aes(x = UMAP1, y = UMAP2),
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  interior_annotation("a", position = c(0.92, 0.93)) +
  xlim(c(-0.25, 1.35)) + ylim(c(-0.25, 2))
```

@@

hex_grid_coloured_scurve2 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve2, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", 
        linewidth=0.2) +
  geom_point(data = umap_scurve_scaled,
           aes(x = UMAP1, y = UMAP2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  xlim(c(-0.25, 1.35)) + ylim(c(-0.25, 2)) +
  interior_annotation("b")

@@

hex_grid_coloured_scurve3 <-  ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve3, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", linewidth=0.2) +
  geom_point(data = umap_scurve_scaled,
           aes(x = UMAP1, y = UMAP2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(c(-0.25, 1.35)) + ylim(c(-0.25, 2)) +
  interior_annotation("c")

@@

```{r}
training_data_scurve <- read_rds("data/s_curve/s_curve_training.rds")
```

<!--To generate errors for different total number of bins-->
```{r}
## To initialize number of bins along the x-axis
bin1_vec_scurve <- 2:19 #sqrt(NROW(training_data_scurve)/2)
buffer_vec <- seq(0.05, 0.1, by = 0.01)

@@

<!-- langevitour with pD model for S-curve -->
::: {#fig-scurve-sc-best layout-ncol="4" fig-pos="H"}
![](figures/scurve/tsne_trimesh_layout.png)

![](figures/scurve/sc_tsne_best_1.png)

![](figures/scurve/sc_tsne_best_2.png)

![](figures/scurve/sc_tsne_best_3.png)

Model in \gD{}, on the tSNE layout, and three views of the fit in projections from $7\text{-}D$, for the S-curve data ($(s_1, \ s_2) = (-0.170, \ -0.137)$, $b = 256 \  (16, \ 16)$, $m = 70$, benchmark value to remove low density hexagons is $0.250$, and benchmark value to remove large edges is $0.180$). MSE is $0.0432$. The model closely fits the shape, but it has some twists (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/ZuLdp89qJ6g>).
:::

>>>>

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: 
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Nonlinear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional (high-D) data using non-linear transformation. The methods and parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of high-D distributions. To help evaluate the NLDR we have developed an algorithm to show the 2D NLDR model in the high-D space, viewed with a tour. One can see if the model fits everywhere or better in some subspaces, or completely mismatches the data. It is used to evaluate which 2D layout is the best representation of the high-D distribution and see how different methods may have similar summaries or quirks.

keywords: [high-dimensional data, dimension reduction, hexagon binning, low-dimensional manifold, tour, data vizualization, model in the data space]
keywords-formatted: [high-dimensional data, dimension reduction, hexagon binning, low-dimensional manifold, tour, data vizualization, model in the data space]

@@

sc_ltr_pos <- c(0.08, 0.93)
# sc_xlims <- c(-0.5, 1.47)
# sc_ylims <- c(-0.32, 2.1)
sc_xlims <- c(-0.25, 1.25)
sc_ylims <- c(-0.2, 1)

@@

Three views of the true model in projections from $7\text{-}D$, for the S-curve data. The model closely fits the shape. (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/tYATMiSlCXQ>). 
:::

XXX The points on the "theoretical model" should be equidistant along the S. They are still to bunched in some spots and too far apart in others.

@@

<!--langevitour with pD model for S-curve-->
<!-- ::: {#fig-scurve-sc layout-ncol="4" fig-pos="H"} -->
<!-- ![](figures/scurve/sc_best_trimesh.png) -->

<!-- ![](figures/scurve/sc_1.png) -->

<!-- ![](figures/scurve/sc_2.png) -->

<!-- ![](figures/scurve/sc_3.png) -->

<!-- XXX Model in \gD{}, on the UMAP layout, and three views of the fit in projections from $7\text{-}D$, for the S-curve data. The model closely fits the shape, but it has breaks which means that it does not adequately capture the surface. (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/N6iWEj3aVps>).  -->
<!-- ::: -->

XXX You need to have consistent colouring. In langevitour the points corresponding to observations are purple. They should be purple in the \gD{} view too.

@@

hex_grid_coloured_scurve1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve1, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  geom_point(data = umap_scurve_scaled,
             aes(x = UMAP1, y = UMAP2),
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  interior_annotation("a", position = c(0.92, 0.93)) +
  xlim(c(-0.25, 1.35)) + ylim(c(-0.25, 1.1))
```

@@

hex_grid_coloured_scurve2 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve2, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", 
        linewidth=0.2) +
  geom_point(data = umap_scurve_scaled,
           aes(x = UMAP1, y = UMAP2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  xlim(c(-0.25, 1.35)) + ylim(c(-0.25, 1.1)) +
  interior_annotation("b")

@@

hex_grid_coloured_scurve3 <-  ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve3, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", linewidth=0.2) +
  geom_point(data = umap_scurve_scaled,
           aes(x = UMAP1, y = UMAP2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(c(-0.25, 1.35)) + ylim(c(-0.25, 1.1)) +
  interior_annotation("c")

@@

```{r}
training_data_scurve <- read_rds("data/s_curve/scurve_500.rds")
```

<!--To generate errors for different total number of bins-->
```{r}
#| eval: false
## To initialize number of bins along the x-axis
bin1_vec_scurve <- 2:19 #sqrt(NROW(training_data_scurve)/2)
buffer_vec <- seq(0.05, 0.1, by = 0.01)

@@

<!--langevitour with pD model for S-curve--> 
<!-- ::: {#fig-scurve-sc-best layout-ncol="4" fig-pos="H"} -->
<!-- ![](figures/scurve/sc_best_trimap_trimesh.png) -->

<!-- ![](figures/scurve/sc_trimap_best_1.png) -->

<!-- ![](figures/scurve/sc_trimap_best_2.png) -->

<!-- ![](figures/scurve/sc_trimap_best_3.png) -->

<!-- Model in \gD{}, on the TriMAP layout, and three views of the fit in projections from $7\text{-}D$, for the S-curve data. The model closely fits the shape, but it has twists. (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/zEz2MUCtADM>).  -->
<!-- ::: -->
